---
title: "Exploring the NOAA Storm Database"
author: "Berj Dekramanjian"
date: "7/29/2021"
output: html_document
---

## Synopsis:

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.


This study will involve exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of  weather events in the United States, starting in the year 1950 and end in November 2011.

The study will analyse which types of events  are most harmful with respect to population health, and have the greatest economic consequences.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing 

The data has been downloaded from the link provided on: <https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2>

It was unzipped using 7-zip and saved to the directory utilized by the project

```{r data processing}
data <- read.csv("StormData.csv")
str(data)
summary(data)
library(ggplot2)
library(dplyr)
library(tidyr)
library(wesanderson)
```


## Extracting variables of Interest  and Transforming Extracted Variables

```{r echo = FALSE}
data1 <- data %>%  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

```
###combine events into catagories
```{r}
data1$EVENT <- "OTHER"
data1$EVENT[grep("HAIL", data1$EVTYPE, ignore.case = TRUE)] <- "HAIL"
data1$EVENT[grep("HEAT", data1$EVTYPE, ignore.case = TRUE)] <- "HEAT"
data1$EVENT[grep("FLOOD", data1$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
data1$EVENT[grep("WIND", data1$EVTYPE, ignore.case = TRUE)] <- "WIND"
data1$EVENT[grep("STORM", data1$EVTYPE, ignore.case = TRUE)] <- "STORM"
data1$EVENT[grep("SNOW", data1$EVTYPE, ignore.case = TRUE)] <- "SNOW"
data1$EVENT[grep("TORNADO", data1$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
data1$EVENT[grep("WINTER", data1$EVTYPE, ignore.case = TRUE)] <- "WINTER"
data1$EVENT[grep("RAIN", data1$EVTYPE, ignore.case = TRUE)] <- "RAIN"
```
### listing the transformed event types 
```{r}
sort(table(data1$EVENT), decreasing = TRUE)
```
###sort costs based on the XP given
```{r}
data1$PROPDMGEXP <- as.character(data1$PROPDMGEXP)
data1$PROPDMGEXP[!grepl("K|M|B", data1$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
data1$PROPDMGEXP[grep("K", data1$PROPDMGEXP, ignore.case = TRUE)] <- "3"
data1$PROPDMGEXP[grep("M", data1$PROPDMGEXP, ignore.case = TRUE)] <- "6"
data1$PROPDMGEXP[grep("B", data1$PROPDMGEXP, ignore.case = TRUE)] <- "9"
data1$PROPDMGEXP <- as.numeric(as.character(data1$PROPDMGEXP))
data1$property.damage <- data1$PROPDMG * 10^data1$PROPDMGEXP

data1$CROPDMGEXP <- as.character(data1$CROPDMGEXP)
data1$CROPDMGEXP[!grepl("K|M|B", data1$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
data1$CROPDMGEXP[grep("K", data1$CROPDMGEXP, ignore.case = TRUE)] <- "3"
data1$CROPDMGEXP[grep("M", data1$CROPDMGEXP, ignore.case = TRUE)] <- "6"
data1$CROPDMGEXP[grep("B", data1$CROPDMGEXP, ignore.case = TRUE)] <- "9"
data1$CROPDMGEXP <- as.numeric(as.character(data1$CROPDMGEXP))
data1$crop.damage <- data1$CROPDMG * 10^data1$CROPDMGEXP
```

## Analysis 

###Aggregating Impact on Health

In order to track the damage caused to human health and the economy by various types of events, we aggregate total injuries and fatalities throughout the years.

```{r impact on health}
AggHumanCost <- data1 %>%  
  group_by(EVENT)  %>% 
  summarise(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
  pivot_longer(-EVENT, names_to = "type", values_to = "amount" )

```

###Aggregating Impact on the Economy

In order to track the damage caused to the economy by various types of events, we aggregate total costs throughout the years and also calculate the average cost per incident of each type of event.

```{r impact on Economy}
AggEconCost <- data1 %>%  
  group_by(EVENT)  %>% 
  summarise(LostProperty = sum(property.damage), LostCrop = sum(crop.damage)) %>% 
  pivot_longer(-EVENT, names_to = "type", values_to = "amount" )

AvgEconCost <- data1 %>%  
  group_by(EVENT)  %>% 
  summarise(LostProperty = mean(property.damage), LostCrop = mean(crop.damage)) %>% 
  pivot_longer(-EVENT, names_to = "type", values_to = "amount" )
```


##Results
###Visulising  harm of various events on human health

The below figure shows the aggregated number of Fatalities and injuries caused by various types of events over the years

``` {r}
ggplot(AggHumanCost, aes(x = EVENT, y = amount, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total number of health impact") +
  ggtitle("Weather event types impact on public health") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette="Set1")
```

It is clear from the data that Tornados have throughout the years have had the most significant impact on human health

###Visulising  harm of various events on the Economy
The below figure shows the aggregated cost of property and crop damages caused by various types of events over the years

``` {r}
ggplot(AggEconCost, aes(x = EVENT, y = amount, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total Cost of Economic Damages") +
  ggtitle("Weather event types impact on Economy") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette="YlGnBu")
```


It is clear from the data that Floods have throughout the years have had the most significant impact on economy when combining costs of lost crops and property



